---
title: "Effect of Income Inequality on Environment Performance Index"
author: "Raju Adhikari and Mario Rodriguez"
date: "December 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Commands
```{r commands, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(WDI)
library(countrycode)
library(rio)
library(stargazer)
library(ggplot2)
library(tidyr)
library(plotly)
library(plm)
library(googleVis)

possible_dir <- c('C:/RajuPC/CollaborativeSSDA/Assignments/Assignment4', '/Users/mariorodriguez/Desktop/Assignment4')
repmis::set_valid_wd(possible_dir)

Combined <- read.csv('EPIGINI.csv')

Combined2012 <- read.csv('EPIGINI2012.csv')

Combined2010 <- read.csv('EPIGINI2010.csv')

Combined2016 <- read.csv('EPIGINI2016.csv')

GINImap <- gvisGeoChart(Combined2010, locationvar = 'iso2c',
                      colorvar = 'GINI',
                      options = list(
                          colors = "['lightblue', 'blue']"
                          ))
EPImap <- gvisGeoChart(Combined2010, locationvar = 'iso2c',
                      colorvar = 'EPI.2010',
                      options = list(
                          colors = "['red', 'yellow']"
                          ))

labels <- c('Gini Coefficient', 'GDP Per Capita (PPP)', '(Intecept)')

m1 <- lm(EPI.2012 ~ GINI, Combined2012)

m2 <- lm(EPI.2012 ~ GINI + GDPPerCapPPP, Combined2012)

m3 <- lm(EPIValue ~ GiniCoeff + GDPperCapPPP, Combined)

m4 <- lm(EPI ~ Gini + GDPperCapPPP + PopulationDens + UrbanPop, Combined2016)
summary(m4)

m5 <- plm(EPIValue ~ GiniCoeff + GDPperCapPPP, data=Combined, index= c('country', 'year'), model = 'within')

m6 <- plm(EPIValue ~ GiniCoeff + GDPperCapPPP, data=Combined, index= c('country', 'year'), model = 'random')

```


## Abstract

Abstract goes here.

## Introduction

Introduction goes here.

## Methodology

Methodology goes here.

## Literature Review

Literature review goes here.

## Operationalization

Operationalization goes here.

## Statistical Analysis

Statistical Analysis goes here.

## Robustness

Robustness tests go here.

## Final Analysis

Final Analysis goes here.

## Conclusion

Conclusion goes here.



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
