---
title: "Income Inequality and Environmental Performance"
author: "Raju Adhikari and Mario Rodriguez"
date: "1/12/2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(WDI)
library(countrycode)
library(rio)
library(stargazer)
library(ggplot2)
library(tidyr)
library(plotly)
library(plm)
library(googleVis)

possible_dir <- c('C:/RajuPC/CollaborativeSSDA/Assignments/Assignment4', '/Users/mariorodriguez/Desktop/Assignment4')
repmis::set_valid_wd(possible_dir)

Combined <- read.csv('EPIGINI.csv')

Combined2012 <- read.csv('EPIGINI2012.csv')

GINIMAP2012 <- Combined2012[, c('iso2c','GINI')]

GINImap <- gvisGeoChart(GINIMAP2012, locationvar = 'iso2c',
                      colorvar = 'GINI',
                      options = list(
                          colors = "['#fff7bc', '#d95f0e']"
                          ))
```

## Table of contents

- Research Question
- Theoretical Basis
- Operationalization
- Plots, Maps and other Descriptive Statistics
- Regression and Other Inferential Statistics
- Conclusion

## Research question

Unrefined Question:

Does income inequality have an effect on the environment?

Refined Question:

Does income inequality of a country affect its Environment Performance Index? If so, How?

## Theoretical premise

- Carbon emission and equal countries
- Economic Development and unequal countries
- Environment Performance Index

## Operationalization

- Explanatory Variable: 
    - Gini coefficient (measure of income inequality)
    - Source: World Bank's API
- Dependent Variable: 
    - EPI (measure of environment performance)
    - Source: Yale Center for Environmental Law and Policy
- Data Cleaning:
    - Excel File, Yearly data separated by sheets 
    - Columns included items used to calculate EPI
    - Merge the data-sets
- Control Variables:
    - GDP per capita

## GINI Map {.flexbox .vcenter}

```{r descriptive 2, results='asis'}
print(GINImap, tag = 'chart')
```

## Plot  {.flexbox .vcenter}

```{r descriptive, echo = FALSE, results='asis', message = FALSE, warning = FALSE}
plot_ly(Combined2012, x = ~EPI.2012, y = ~GINI,
        mode = 'markers')
```

## Inferential statistics

## Robustness

```{r robust}
plot(pressure)
```

